Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-06 Manuscript/manuscript.Rmd
intro edits

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect genomic regulation [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given the high degree of conservation between cell types and species, it has been proposed that TADs represent a fundamental unit of organization, giving the genome its architecture [@rocha2015breaking; @dixon2012topological]. Likewise, Evidence suggests that the boundaries that delineate TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes [@symmons2014functional]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d; @symmons2014functional] [??? This sentence and reference belong to the previous paragraph, think and move; !!!combined and moved]. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [dixon2015chromatin; @di2017novo] [??? REFS? Carefully select from my paper notes; !!!added refs]. It has been established that the zinc finger protein, CTCF, is associated with TAD boundaries and is an architectural protein responsible for the organization of the genome within the nucleus [@ghirlando2016ctcf; @ong2014ctcf]. CTCF along with other transcription factors, including SMC1, SMC3, and RAD21 <!--have been shown to -->colocalize <!--together [??? Note removal of "water" words]-->near TAD boundaries, leading to the recruitment of a cohesin ring [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Other binding proteins have been shown to mediate enhancer-promoter interactions, including the transcription factors YY1 and ZNF143, thereby influencing the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. However, the growing number of annotated genomic features and the methods used to identify the ones associated with TAD boundaries still leaves much to be desired. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it.]

@@

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between functional annotations and regions that include TAD boundaries. However, <!--this is a multi-faceted probelem. [??? "WATER"!]-->many key characteristics of 3D genomic data remains unexplored, potentially limiting performance of predictive modeling. <!--including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.-->Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced ratio of positive/negative classes usef to train the models. The imbalance ratio is defined as the ratio of minority classes (regions with TAD boundaries) to majority classes (regions without TAD boundaries) [@Orriols-Puig2008]. [??? You confused me a lot with this] Such heavily imbalanced datasets <!--are known to [??? WATER!] -->adversely impact the performance of the classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, it remains unclear how the increased resolution of Hi-C data affect the performance of predictive modeling. Lastly, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored.

@@

<!--Secondly, it is important to determine the optimal representation of the relationship between features in the model and genomic regions. This is known as feature engineering. Typical approaches include overlap counts [@hong2017computational] or percents [@mourad2016computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e.g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin. Here, using percents offers a more granular approach, but it is unclear how it performs when compared to counts. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. Moreover, while direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving forces of genome regulation [??? REFS]. These relationships can be quantified using distances (in basepairs) from regions on the linear genome to regions that define functional genomic annotations. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. The use of distances have largely been ignored, yet have the potential to offer the most accurate spatial representation of which genomic features are associated with TAD boundaries.-->

Lastly, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from previous studies what resolution contact matrices were used when retrieving TAD boundaries, and how other resolutions would affect their results. A lower resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances. To date, no study has compared the results of their predictive models built across multiple data resolutions.

Furthermore... [??? You never describe the other data property - various levels of relationships]

Data level solutions mostly involve re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. [??? So, if the problem is solved, why do you bother with class imbalance at all?]

Evaluation level solutions include choosing suitable performance metrics for evaluating models. It has been shown that common metrics such as accuracy and AUROC (area under the receiver operating characteristic curve) are insensitive to class imbalance [@prati2011survey; @wei2013role]. A better measure to consider would be a composite metric involving all quadrants of the confusion matrix such as Matthews Correlation Coefficient (MCC). The suitability of MCC when evaluating models in imbalanced settings has been studied previously [@boughorbel2017optimal; @brodersen2010balanced]. Defined, the MCC is the correlation coefficient between the observed and predicted binary classifications. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction. [??? So, if the problem is solved, why do you bother with it at all?]

[??? Carefully think on the "why" questions above. This paragraph answers it, connect it]
In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. The class imbalance issue was not addressed. Even worse, when performing cross-validation, [@mourad2016computational] used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance.

[??? Carefully think on the "why" questions above. This paragraph answers it, connect it]
A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases [?? Need to identify what you mean by "total number of cases." Is it just true positive divided by total number of possible boundaries ??]. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes [@saito2015precision]. Another threshold-specific measure to consider is a composite metric involving all quadrants of the confusion matrix, such as Matthews Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten] [?? MCC actually has all of the same components as AUC. AUC is a function of sensitivity and specificity which combined contain all quandrants ??]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

<--
Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role]. 
-->

[Paragraph about disconnect between Hi-C and epigenomic data] Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?](Zhang et al., “HiCPlus.”, Carron et al., “Boost-HiC.”) and technological [@Hsieh:2019aa][@Ma:1998aa][@Belaghzal2017] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).

In this study, we developed a novel ensemble framework aimed at systematically comparing the performances of random forest classification models in order to determine the set of data level characteristics necessary for optimal prediction of TAD boundaries. We evaluated performances of Hi-C data for both the GM12878 and K562 cell line obtained at four different Hi-C resolutions including 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each cell line and resolution, we compared three feature engineering techniques: counts, percents, and distances. Additionally, we compared three re-sampling techniques: random under-sampling, random-over-sampling, and a weighted combination of both referred to as Synthetic Minority Over-sampling Technique (SMOTE).

[???Include concluding paragraph]

>>>>

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome <!--may affect genomic regulation--> [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking; @dixon2012topological]. Likewise, evidence suggests that the boundaries that delineate TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes [@symmons2014functional]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d; @symmons2014functional] [??? This sentence and reference belong to the previous paragraph, think and move; !!!combined and moved]. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

Distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [@dixon2015chromatin; @di2017novo] [??? REFS? Carefully select from my paper notes; !!!added refs]. Recent research has identified key functional genomic annotations enriched at boundaries. Namely, the zinc finger protein, CTCF, has been shown to be associated with TAD boundaries [@ghirlando2016ctcf]. Other transcription factors including SMC1, SMC3, and RAD21 have been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Additional binding proteins, known to mediate enhancer-promoter interactions including YY1 and ZNF143, have been shown to influence the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. <!--However, there is a growing number of publically available annotated genomic features, and it is unclear if current methods are robust enough to identify all of them. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it. !!!editted]-->

@@

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and regions that include TAD boundaries. However, <!--this is a multi-faceted probelem. [??? "WATER"!]-->many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. <!--including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.-->Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. The imbalance ratio is defined as the ratio of minority classes (regions with TAD boundaries) to majority classes (regions without TAD boundaries) [@Orriols-Puig2008]. [??? You confused me a lot with this] Such heavily imbalanced datasets <!--are known to [??? WATER!] -->adversely impact the performance of the classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how the increased resolution of Hi-C data affect the performance of predictive modeling.

@@

Recently, an L1-regularized multiple logistic regression model was developed to predict TAD boundaries using transcription factor binding sites on GM12878 [@mourad2016computational]. Based on a 1 kb binned genomic data matrix and 20 kb flanked TAD boundaries, the model was trained to identify the set of transcription factors most influential of TAD boundaries. Several problems exist with the study. First, the class imbalance issue was not addressed. Secondly, many transcription factors are known to colocalize, and therefore exhibit high correlations. If a group of predictors are highly correlated among themselves, L1-regularization tends to pick only one of them and will shrink the others to zero [@zou2005regularization].  Thirdly, [@mourad2016computational] reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. It has been shown that AUC is insensitive to class imbalance, creating artificially inflated values inlfuenced by the majority class [@prati2011survey; @wei2013role].
